---
title: "Learning Report Analysis Outputs"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-packages}
pacman::p_load(
  "tidyverse", 
  "lubridate", 
  "glue"
)
```


## Introduction

### Purpose

### Load data

```{r load-data}
set.seed(489)

dat <- read_rds("data/tables-extract.rds") %>% 
  mutate(
    id = row_number(), 
    cohort = c("Cohort 1", "Cohort 2") %>% sample(n(), replace = TRUE)
  )
dat %>% skimr::skim()
```

`clean_data` generates a single data frame that contains all the information presented in the report analysis. 

## Analysis

### Summarize a single boolean value

```{r summary-boolean-one}
# Summarize the racialized variable
tab_racialized <- dat %>% 
  
  filter(!is.na(demo_racialized)) %>% 
  
  # Summarize the following values
  summarize(
    N = n(), # the sample size
    n = sum(demo_racialized), # the number of TRUEs
    p = mean(demo_racialized) # the proportion of TRUEs
  ) %>% 
  
  mutate(
    pct = scales::percent(p, accuracy = 1), # print percentages
    pct_n = glue("{pct} ({n}/{N})") # use glue to finalize values
  )

tab_racialized %>% 
  
  # select relevant outputs
  select(pct_n) %>% 
  
  # convert table to markdown
  knitr::kable()
```

### Summarize a number of boolean variables

```{r summary-boolean-many}
# Summarize all boolean demographic variables in the data
tab_demo_bool <- dat %>% 
  
  # pivot boolean variables to long format
  pivot_longer(
    cols = c(matches("demo")), 
    names_to = "variable", 
    values_to = "value"
  ) %>% 
  
  # filter out rows with na values
  filter(!is.na(value)) %>% 
  
  # group by variable name
  group_by(variable) %>% 
  
  # Summarize the following values
  summarize(
    N = n(), # the sample size
    n = sum(value), # the number of TRUEs
    p = mean(value) # the proportion of TRUEs
  ) %>% 
  
  mutate(
    pct = scales::percent(p, accuracy = 1), # print percentages
    pct_n = glue("{pct} ({n}/{N})") # use glue to finalize values
  )

tab_demo_bool %>% 
  
  # select relevant outputs
  select(variable, pct_n) %>% 
  
  # convert table to markdown
  knitr::kable()
```

### Summarize a number of boolean variables by a categorical variable

```{r summary-boolean-many-by-cohort}
tab_demo_by_cohort <- dat %>% 
  
  pivot_longer(
    cols = c(matches("demo")), 
    names_to = "variable", 
    values_to = "value"
  ) %>% 
  
  filter(!is.na(value)) %>% 
  
  group_by(cohort, variable) %>% 
  
  # Summarize the following values
  summarize(
    N = n(), # the sample size
    n = sum(value), # the number of TRUEs
    p = mean(value) # the proportion of TRUEs
  ) %>% 
  
  mutate(
    pct = scales::percent(p, accuracy = 1), # print percentages
    pct_n = glue("{pct} ({n}/{N})") # use glue to finalize values
  )

tab_demo_by_cohort %>% 
  
  # select relevant outputs
  select(variable, cohort, pct_n) %>% 
  
  # pivot crosstab for readability
  pivot_wider(
    names_from = cohort, 
    values_from = pct_n
  ) %>% 
  
  # convert table to markdown
  knitr::kable()
```

### Add a "Total" column

```{r summary-boolean-many-by-cohort-total}
tab_demo_by_cohort_with_total <- dat %>% 
  
  bind_rows(dat %>% mutate(cohort = "Total")) %>% 
  
  pivot_longer(
    cols = c(matches("demo")), 
    names_to = "variable", 
    values_to = "value"
  ) %>% 
  
  filter(!is.na(value)) %>% 
  
  group_by(cohort, variable) %>% 
  
  # Summarize the following values
  summarize(
    N = n(), # the sample size
    n = sum(value), # the number of TRUEs
    p = mean(value) # the proportion of TRUEs
  ) %>% 
  
  mutate(
    pct = scales::percent(p, accuracy = 1), # print percentages
    pct_n = glue("{pct} ({n}/{N})") # use glue to finalize values
  )

tab_demo_by_cohort_with_total %>% 
  
  # select relevant outputs
  select(variable, cohort, pct_n) %>% 
  
  pivot_wider(
    names_from = cohort, 
    values_from = pct_n
  ) %>% 
  
  # convert table to markdown
  knitr::kable()
```

### Summarize a categorical variable

```{r summary-categorical-one}
tab_educ <- dat %>% 
  
  filter(!is.na(educ_highest)) %>% 
  
  group_by(educ_highest, .drop = FALSE) %>% 
  
  summarize(
    n = n()
  ) %>% 
  
  mutate(
    N = sum(n), 
    p = n / N, 
    pct = scales::percent(p, accuracy = 1), 
    pct_n = glue("{pct} ({n}/{N})")
  )

tab_educ %>% 
  select(
    educ_highest, 
    pct_n
  ) %>% 
  
  knitr::kable()
```

### Summarize a categorical variable by another categorical variable

```{r summary-categorical-two}
tab_educ_by_cohort <- dat %>% 
  
  bind_rows(dat %>% mutate(cohort = "Total")) %>% 
  
  filter(!is.na(educ_highest)) %>% 
  
  group_by(
    cohort, educ_highest, .drop = FALSE
  ) %>% 
  
  summarize(
    n = n()
  ) %>% 
  
  group_by(
    cohort
  ) %>% 
  
  mutate(
    N = sum(n), 
    p = n / N, 
    pct = scales::percent(p, accuracy = 1), 
    pct_n = glue::glue("{pct} ({n}/{N})")
  )

tab_educ_by_cohort %>% 
  select(
    cohort, educ_highest, pct_n
  ) %>% 
  
  pivot_wider(
    names_from = cohort, 
    values_from = pct_n
  ) %>% 
  
  knitr::kable()

```


